---
title: "cleaning_aus_rls"
author: "Sarah Willington"
date: "2026-05-21"
output: html_document
---

script to clean and explore raw rls data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)


```

```{r libraries}
library(tidyverse)
library(here)
library(patchwork)
library(tidylog)
library(gt)
library(sf)
library(rnaturalearth)
library(rnaturalearthhires)
library(paletteer)
library(ggtext)
```


```{r set project directory path}

rm(list = ls())
project_dir <- here::here()
```


## read data

### initial filtering. 

1. Refine to only include the temperate SE Aus regions (NSW, VIC, TAS)

2. Filter to only include method 1 surveys

3. Filter to include fish only

4. remove Actinopterygii spp. under species_name

5. Remove size class = 0

7. Assign to populations (1 degree lat lon boxes)

8. Keep species × population × year combinations with at least 50 individuals

9. spread across at least 5 size classes.

10. Then keep only species × population combinations with at least 10 years of data.

```{r load data and filter}

# d <- read_csv(file.path(project_dir, "0_data", "01_data_raw", "rls_aus_raw.csv"), skip = 69)
# col_id <- d[1:69,]
# write_csv(col_id, "0_data/01_data_raw/col_ids.csv")

# read in data excluding the meta data
df <- read_csv(
  file.path(project_dir, "0_data", "01_data_raw", "rls_aus_raw.csv"),
  skip = 70,
  quote = "",
  show_col_types = FALSE) |> 
  rename_with(~ str_remove_all(.x, '"')) |> 
  mutate(across(where(is.character), ~ str_remove_all(.x, '"')))

# filter for areas of interest and for fish only. 
# remove size class 0 
# create year col
# remove unwanted cols
# make necessary cols numeric
df2 <- df |> 
  filter(area == "New South Wales" | area == "Victoria" | area == "Tasmania") |> # SE aus temperate regions
  filter(method == 1) |> 
  filter(class == "Actinopteri" | class == "Actinopterygii" | class == "Teleostei") |> 
  filter(species_name!="Actinopterygii spp.") |> 
  filter(size_class!=0) |> 
  separate(survey_date, c('year', 'month', 'day'), sep='-', remove=TRUE) |> 
  select(-c(FID, country, realm, block, reporting_name, geom, visibility, survey_latitude, survey_longitude)) |>  
  mutate(latitude = as.numeric(latitude), 
         longitude = as.numeric(longitude), 
         year = as.numeric(year), 
         month = as.numeric(month), 
         day = as.numeric(day), 
         depth = as.numeric(depth),
         size_class = as.numeric(size_class), 
         total = as.numeric(total)) |> 
  rename(species = species_name) 

# assign populations
lon_breaks <- seq(135.5, 160.5, by = 1)
lat_breaks <- seq(-45.5, -10.5, by = 1)

lon_labels <- lon_breaks[-length(lon_breaks)] + 0.5
lat_labels <- lat_breaks[-length(lat_breaks)] + 0.5

df_pop <- df2 |>
  mutate(
    pop_lon = cut(
      longitude,
      breaks = lon_breaks,
      labels = lon_labels,
      include.lowest = TRUE),
    pop_lat = cut(
      latitude,
      breaks = lat_breaks,
      labels = lat_labels,
      include.lowest = TRUE),
    pop_lon = as.numeric(as.character(pop_lon)),
    pop_lat = as.numeric(as.character(pop_lat)),
    population = paste(pop_lat, pop_lon, sep = "_"))|> 
   mutate(grp_id = row_number())

# Keep species × population × year combinations with at least 50 individuals
# spread across at least 5 size classes.
# Then keep only species × population combinations with at least 10 years of data.
df_filtered <- df_pop |> 
  group_by(species, population, year) |> 
  mutate(n_obs_pop = sum(total),
         n_size_class = n_distinct(size_class)) |> 
  filter(n_obs_pop>=50,
         n_size_class>=5) |> 
  ungroup() |>
  group_by(species, population) |>
  mutate(n_year = n_distinct(year)) |>
  filter(n_year>=10) |> 
  ungroup()

```

```{r summary tables}
# create summary tables to explore data
species_summary <- df_filtered |> 
  arrange(desc(total)) |>
  group_by(species) |>
  summarise(n_obs = sum(total),
            n_year = n_distinct(year),
            n_size_class = n_distinct(size_class),
            n_pop = n_distinct(population)) |> 
  arrange(desc(n_obs)) |> 
  gt() |> 
  cols_label(species = "Species") |> 
  tab_options(table.font.size = px(12))

species_summary

# create df of obs per size class per species
size_class_counts <- df_filtered |>
  group_by(species, size_class) |>
  summarise(n = sum(total), .groups = "drop") |>
  pivot_wider(
    names_from = size_class,
    values_from = n,
    values_fill = 0,
    names_prefix = "size_"
  )

# summary table of species x size classes
species_summary <- df_filtered |> 
  group_by(species) |>
  summarise(
    n_obs = sum(total),
    n_year = n_distinct(year),
    n_size_class = n_distinct(size_class),
    n_sites = n_distinct(site_code),
    .groups = "drop"
  ) |>
  left_join(size_class_counts, by = "species") |>
  arrange(desc(n_obs)) |> 
  gt() |> 
  cols_label(species = "Species") |> 
  tab_options(table.font.size = px(12))
species_summary
  
```

```{r filter species}
# filter species
# remove all pelagic species
# remove species that are heavily fished

df_filtered <- df_filtered |> 
  filter(species!="Trachurus novaezelandiae"& # pelagic. yellowtail scad
           #species!="Trachinops caudimaculatus"& # southern hulafish. Max length 15 cm TL
           #species!="Parma microlepis"& # white-ear. Max length 20 cm TL
           species!="Pagrus auratus"& # snapper. fished
           species!="Nemadactylus douglasii"& # grey morwong. fished
           species!="Trachurus declivis"& # pelagic. greenback horse mackerel
           species!="Acanthopagrus australis" &# yellowfin bream. fished
           species!= "Rhabdosargus sarba"& # tarwhile, seabream. fished
           species!= "Scomber australasicus"& # blue mack. pelagic
           #species!="Scorpis lineolata"& # silver sweep. rec fish
           #species!="Scorpis aequipinnis"& # sea sweep. rec fish
           #species!="Notolabrus tetricus"& # blue throat wrasse. fished
           #species!="Dinolestes lewini"& # long fin pike. rec fish
           #species!="Notolabrus fucicola"& # purple wrasse. fished
           #species!="Girella tricuspidata"& # luderick. rec fish
           #species!="Chirodactylus spectabilis"& # banded morwong. fished
           #species!="Kyphosus sydneyanus"& # silver drummer. rec fish
           #species!="Girella elevata"& # rec fish
           #species!="Latridopsis forsteri"& # bastard trumpeter. fished
           species!="Pseudocaranx dentex") # silver trevally

d_sum <- df_filtered |> group_by(species) |> 
  summarise(n_obs = sum(total), 
            n_year = n_distinct(year),
            n_size_class = n_distinct(size_class),
            n_sites = n_distinct(site_code),
            area   = paste(sort(unique(area)), collapse = ", "),)
  
#write_csv(d_sum, "0_data/01_data_raw/species_list.csv")
```

```{r visualise data plots}

d_points <- df_filtered |> 
  select(area, species, year, size_class, total) 

d_lines <- df_filtered |> 
  group_by(area, species, year, size_class) |> 
  summarise(
    n_obs_size = sum(total, na.rm = TRUE),
    .groups = "drop") 

plot_groups <- d_points |> 
  distinct(species, area)

pwalk(plot_groups, function(species, area) {
  
  point_data <- d_points |> 
    filter(species == !!species, area == !!area)
  
  line_data <- d_lines |> 
    filter(species == !!species, area == !!area)
  
   # scaling values for this species x area plot
  point_max <- max(point_data$total, na.rm = TRUE)
  line_max  <- max(line_data$n_obs_size, na.rm = TRUE)
  
  scale_factor <- point_max / line_max
  
  p <- ggplot(point_data, aes(x = size_class)) +
    geom_point(aes(y = total), size = 0.7, alpha = 0.7) +
    geom_line( data = line_data,
              aes(y = n_obs_size* scale_factor, group = 1),
              colour = "grey40",
              linewidth = 0.4) +
    scale_y_continuous(
      name = "Total",
      sec.axis = sec_axis(
        ~ . / scale_factor,
        name = "Total across observations")) +
    labs(title = paste(species, area, sep = " | "),
      x = "Size class",
      y = "Total") +
    facet_wrap(~ year, scales = "free_y") +
    theme_bw() +
    theme(axis.text = element_text(size = 7))
  
  file_name <- paste(species, area, sep = "_") |> 
    str_replace_all("[^A-Za-z0-9]+", "_")
  
  ggsave(
    filename = paste0("02_pipeline/01_data_cleaning/plots_species_area/", file_name, ".png"),
    plot = p,
    width = 9,
    height = 6,
    dpi = 300
  )
})


```

```{r}
plots_tbl <- df_filtered |> 
  mutate(size_class = factor(size_class)) |> 
  group_nest(species, area) |> 
  mutate(
    plot = pmap(
      list(data, species, area),
      ~ ggplot(..1, aes(x = size_class, y = total)) +
        geom_point() +
        labs(
          title = paste(..2, ..3, sep = " - "),
          x = "Size class",
          y = "Total"
        ) +
        facet_wrap(~ year, scales = "free_y") +
        theme_bw()
    )
  )

plots_tbl |> 
  filter(species == "Scorpis lineolata", area == "New South Wales") |> 
  pull(plot) |> 
  pluck(1)

```

```{r manually filter outliers}
df_filtered2 <- df_filtered |> 
  filter(!(species == "Acanthaluteres vittiger" & size_class >= 50)) |> 
  filter(!(species == "Achoerodus viridis" & total >100)) |> 
  filter(!(species == "Achoerodus viridis" & size_class >=150 )) |> 
  filter(!(species == "Anampses elegans" & size_class > 40 )) |> 
  filter(!(species == "Aplodactylus lophodon" & size_class>50)) |> 
  filter(!(species == "Atypichthys strigatus" & total > 1000 )) |>
  filter(!(species == "Atypichthys strigatus" & size_class >35 )) |> 
  filter(!(species == "Austrolabrus maculatus" & size_class >20 )) |> 
  filter(!(species == "Caesioperca lepidoptera" & total>= 1000)) |> 
  filter(!(species == "Caesioperca lepidoptera" & size_class>35)) |> 
  filter(!(species == "Chaetodon tricinctus" & total>= 500)) |> 
  filter(!(species == "Chlorurus dordidus" & size_class>60)) |>
  filter(!(species == "Chromis hypsilepis" & size_class>25)) |>
  filter(!(species == "Chromis hypsilepis" & total >=2000)) |>
  filter(!(species == "Coris bulbifrons" & size_class > 80)) |> 
  filter(!(species == "Coris picta" & size_class >= 40)) |> 
  filter(!(species == "Dinolestes lewini" & size_class>70)) |> 
  filter(!(species == "Dotalabrus aurantiacus" & size_class>20)) |> 
  filter(!(species == "Enoplosus armatus" & size_class > 40)) |> 
  filter(!(species == "Eupetrichthys angustipes" & size_class > 25)) |> 
  filter(!(species == "Gerres subfasciatus" & size_class > 30)) |>
  filter(!(species == "Girella zebra" & size_class>60)) |> 
  filter(!(species == "Halichoeres nebulosus" & size_class > 15)) |> 
  filter(!(species == "Hypoplectrodes maccullochi" & size_class > 30)) |> 
  filter(!(species == "Labroides dimidiatus" & size_class > 20)) |> 
  filter(!(species == "Mecaenichthys immaculatus" & size_class > 20)) |>
  filter(!(species == "Meuschenia freycineti" & size_class>70)) |> 
  filter(!(species == "Meuschenia hippocrepis" & size_class > 60)) |> 
  filter(!(species == "Morwong fuscus" & size_class >75)) |> 
  filter(!(species == "Morwong fuscus" & total>300)) |> 
  filter(!(species == "Neoglyphidodon polyacanthus" & size_class > 20)) |> 
  filter(!(species == "Neoodax balteatus" & size_class > 25)) |> 
  filter(!(species == "Notolabrus fucicola" & size_class>60)) |> 
  filter(!(species == "Notolabrus gymnogenis" & size_class>60)) |> 
  filter(!(species == "Notolabrus tetricus" & size_class > 70)) |> 
  filter(!(species == "Olisthops cyanomelas" & size_class>=75)) |> 
  filter(!(species == "Ophthalmolepis lineolatus" & size_class>60)) |> 
  filter(!(species == "Ophthalmolepis lineolatus" & total >75)) |>
  filter(!(species == "Ostorhinchus limenus" & size_class>20)) |>
  filter(!(species == "Ostorhinchus norfolcensis" & size_class>20)) |>
  filter(!(species == "Ostorhinchus norfolcensis" & total >=1000)) |>
  filter(!(species == "Parma microlepis" & size_class>30)) |> 
  filter(!(species == "Parma microlepis" & total >=80)) |> 
  filter(!(species == "Parma unifasciata" & size_class>25)) |> 
  filter(!(species == "Parma victoriae" & size_class> 35)) |> 
  filter(!(species == "Pempheris affinis" & size_class>25)) |> 
  filter(!(species == "Pempheris affinis" & total >=1000)) |>
  filter(!(species == "Pempheris compressa" & total >=1000)) |> 
  filter(!(species == "Pictilabrus laticlavius" & size_class>=40)) |> 
  filter(!(species == "Pictilabrus laticlavius" & total >200)) |> 
  filter(!(species == "Prionurus microlepidotus" & total>=1000)) |> 
  filter(!(species == "Pseudogoniistius nigripes" & size_class>60)) |> 
  filter(!(species == "Pseudogoniistius nigripes" & total>100)) |> 
  filter(!(species == "Pseudolabrus guentheri" & size_class>30 )) |> 
  filter(!(species == "Pseudolabrus luculentus" & size_class>30 )) |> 
  filter(!(species == "Pseudolabrus rubicundus" & size_class>30 )) |> 
  filter(!(species == "Schuettea scalaripinnis" & total>1000 )) |>
  filter(!(species == "Scorpis aequipinnis" & size_class>60)) |> 
  filter(!(species == "Scorpis lineolata" & total>1500)) |> 
  filter(!(species == "Scorpis lineolata" & size_class >40)) |> 
  filter(!(species == "Stegastes fasciolatus" & size_class >20)) |> 
  filter(!(species == "Thalassoma amblycephalum" & size_class >20)) |> 
  filter(!(species == "Thalassoma lunare" & size_class >40)) |> 
  filter(!(species == "Thalassoma lunare" & total >=1000)) |> 
  filter(!(species == "Trachinops caudimaculatus" & size_class>20)) |> 
  filter(!(species == "Trachinops caudimaculatus" & total >4000)) |> 
  filter(!(species == "Trachinops taeniatus" & size_class>20)) |> 
  filter(!(species == "Trachinops taeniatus" & total >=4000)) |> 
  filter(!(species == "Upeneichthys lineatus" & size_class >=40)) 
```


```{r species distribution maps}

# Summarise to one point per site
site_points <- df_filtered2|>
  group_by(species, site_code) |>
  summarise(
    lon = first(longitude),
    lat = first(latitude),
    n_obs = sum(total, na.rm = TRUE),
    .groups = "drop")

# Convert to sf points
site_points_sf <- st_as_sf(
  site_points,
  coords = c("lon", "lat"),
  crs = 4326)

# Australia land
aus <- rnaturalearth::ne_download(
  scale = 10,
  type = "land",
  category = "physical",
  returnclass = "sf") |>
  st_transform(4326)

# Crop to southeastern Australia / Tasmania
aus <- st_crop(
  aus,
  xmin = 135, xmax = 155,
  ymin = -45, ymax = -20)

# state boundaries
aus_states <- rnaturalearth::ne_states(
  country = "Australia",
  returnclass = "sf") |>
  st_transform(4326) |>
  st_crop(
    xmin = 135, xmax = 155,
    ymin = -45, ymax = -20 )

# Set plot limits from your points, with padding
pad <- 1.5

point_coords <- st_coordinates(site_points_sf)

xlim <- range(point_coords[, "X"], na.rm = TRUE) + c(-pad, pad)
ylim <- range(point_coords[, "Y"], na.rm = TRUE) + c(-pad, pad)

# Plot
species_maps <- site_points_sf |>
  split(site_points_sf$species) |>
  map(~ {
    ggplot() +
      geom_sf(data = aus, fill = "grey85", colour = "grey60", linewidth = 0.2) +
      geom_sf(data = aus_states, fill = NA, colour = "grey65", linewidth = 0.25) +
      geom_sf(
        data = .x,
        aes(size = n_obs),
        colour = "black",
        alpha = 0.8 ) +
      scale_size_continuous(
        name = "No. of obs",
        range = c(0.8, 4) ) +
      coord_sf(
        xlim = xlim,
        ylim = ylim,
        expand = FALSE) +
      scale_y_continuous( breaks = seq(-45.5, -10.5, by = 1),
                          labels = function(x) ifelse(x %in% seq(-45.5, -10.5, by = 2), x, "")) +
      scale_x_continuous(breaks = seq(135.5, 160.5, by = 1),
                         labels = function(x) ifelse(x %in% seq(135.5, 160.5,, by = 4), x, ""))+
      labs(
        title = unique(.x$species),
        x = expression("Longitude ("*degree*"E)"),
        y = expression("Latitude ("*degree*"S)")) +
      theme_bw() +
      theme(legend.position = "right" )
  })

species_maps[[35]]

# facet plot of all species
p_species_maps <- ggplot() +
  geom_sf(data = aus, fill = "grey85", colour = "grey60", linewidth = 0.2) +
  geom_sf(data = aus_states, fill = NA, colour = "grey65", linewidth = 0.25) +
  geom_sf(
    data = site_points_sf,
    size = 0.5,
    colour = "black",
    alpha = 0.8) +
  coord_sf(
    xlim = xlim,
    ylim = ylim,
    expand = FALSE) +
  facet_wrap(~ species,
             labeller = labeller(species = label_wrap_gen(width = 10)), 
             ncol = 9) +
  scale_y_continuous( breaks = seq(-45.5, -10.5, by = 1),
                          labels = function(x) ifelse(seq_along(x) %% 5 == 1, x, "")) +
      scale_x_continuous(breaks = seq(135.5, 160.5, by = 1),
                         labels = function(x) ifelse(seq_along(x) %% 6 == 1, x, ""))+
  labs(
    x = expression("Longitude ("*degree*"E)"),
    y = expression("Latitude ("*degree*"S)")) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 15),
    axis.ticks = element_line(linewidth = 0.2),
    axis.text = element_text(size = 8), 
    axis.text.x = element_text(angle = 90, hjust = 1),
    strip.text = element_text(size = 9, face = "italic"))

p_species_maps

ggsave(
  "02_pipeline/01_data_cleaning/species_site_maps2.png",
  p_species_maps,
  width = 32,
  height = 42,
  units = "cm",
  dpi = 600)

```

```{r map showing populations}
map_points <- df_filtered2 |> 
  select(species, site_code, total, population, latitude, longitude) |> 
  distinct(species, site_code, .keep_all = TRUE)

ggplot() +
  geom_sf(data = aus, fill = "grey85", colour = "grey60", linewidth = 0.2) +
  geom_sf(data = aus_states, fill = NA, colour = "grey65", linewidth = 0.25) +
  geom_point(
    data = map_points,
    aes(x = longitude, y = latitude, colour = population),
    size = 1
  ) +
  coord_sf(
    xlim = xlim,
    ylim = ylim,
    expand = FALSE,
    default_crs = sf::st_crs(4326)
  ) +
  scale_y_continuous(breaks = seq(-45.5, -10.5, by = 1),
                     labels = function(x) ifelse(seq_along(x) %% 2 == 1, x, "")) +
  scale_x_continuous(breaks = seq(135.5, 160.5, by = 1),
                     labels = function(x) ifelse(seq_along(x) %% 5 == 1, x, "")) +
  labs(
    x = expression("Longitude ("*degree*"E)"),
    y = expression("Latitude ("*degree*"S)")
  ) +
  theme_bw()+
  theme(axis.ticks = element_line(linewidth = 0.2), 
        legend.position = "none")

ggsave(
  "02_pipeline/01_data_cleaning/species_populations.png",
  last_plot(),
  width = 10,
  height = 15,
  units = "cm",
  dpi = 300)
```

```{r annual size dist plots with map}
# set up data 
d_points <- df_filtered2 |> 
  select(area, species, year, size_class, total) 

d_lines <- df_filtered2 |> 
  group_by(area, species, year, size_class) |> 
  summarise(
    n_obs_size = sum(total, na.rm = TRUE),
    .groups = "drop") 

plot_groups <- d_points |> 
  distinct(species, area)

map_points <- df_filtered2 |> 
  select(species, site_code, total, population, latitude, longitude) |> 
  distinct(species, site_code, .keep_all = TRUE)

# loop through species
pwalk(plot_groups, function(species, area) {
  
  point_data <- d_points |> 
    filter(species == !!species, area == !!area)
  
  line_data <- d_lines |> 
    filter(species == !!species, area == !!area)
  
  map_data <- map_points |>
    filter(species == !!species)
  
  point_max <- max(point_data$total, na.rm = TRUE)
  line_max  <- max(line_data$n_obs_size, na.rm = TRUE)
  
  scale_factor <- point_max / line_max
  
  pop_palette <- paletteer_d("MetBrewer::Redon")
  
  p_years <- ggplot(point_data, aes(x = size_class)) +
    geom_point(aes(y = total), size = 0.7, alpha = 0.8, shape = 16) +
    geom_line(
      data = line_data,
      aes(y = n_obs_size * scale_factor, group = 1),
      colour = "darkcyan",
      linewidth = 0.4) +
    scale_y_continuous(
      name = "Total per survey",
      sec.axis = sec_axis( ~ . / scale_factor, name = "Cumulative total") ) +
    facet_wrap(~ year, scales = "free_y", ncol = 5) +
    labs(title = paste0("<i>", species, "</i> | ", area),
      x = "Size class" ) +
    theme_bw() +
    theme(plot.title = ggtext::element_markdown(size = 13),
      axis.text = element_text(size = 6),
      strip.text = element_text(size = 8), 
      axis.title = element_text(size = 12), 
      panel.grid.minor = element_blank())
  
  p_map <- ggplot() +
    geom_sf(data = aus, fill = "grey85", colour = "grey60", linewidth = 0.2) +
    geom_sf(data = aus_states, fill = NA, colour = "grey65", linewidth = 0.25) +
    geom_point(
      data = map_data,
      aes(x = longitude, y = latitude, colour = population),
      size = 0.8,
      alpha = 0.8, 
      shape = 16) +
    coord_sf(
      xlim = xlim,
      ylim = ylim,
      expand = FALSE) +
    scale_colour_manual(values = pop_palette) +
    scale_y_continuous(breaks = seq(-45.5, -10.5, by = 1),
                       labels = function(x) ifelse(seq_along(x) %% 4 == 1, x, "")) +
  scale_x_continuous(breaks = seq(135.5, 160.5, by = 1),
                     labels = function(x) ifelse(seq_along(x) %% 5 == 1, x, "")) +
    labs(
       x = expression("Longitude ("*degree*"E)"),
        y = expression("Latitude ("*degree*"S)")) +
    theme_bw() +
    theme(
      axis.ticks = element_line(linewidth = 0.2),
      axis.text = element_text(size = 6),
      axis.title = element_blank(),
      strip.text = element_text(size = 8),
      legend.position = "none")
  
  p <- p_years + p_map +
    plot_layout(widths = c(4, 0.85))
  
  file_name <- paste(species, area, sep = "_") |> 
    str_replace_all("[^A-Za-z0-9]+", "_")
  
  ggsave(
    filename = paste0(
      "02_pipeline/01_data_cleaning/plots_species_area_map/",
      file_name,
      ".png"),
    plot = p,
    width = 12,
    height = 7,
    dpi = 600 )
})
```

```{r species heat maps}
#set up data 
data_pop_year <- df_filtered2 |> 
  group_by(species, population, year) |> 
  summarise(
    n_individuals = sum(total, na.rm = TRUE),
    n_size_classes = n_distinct(size_class),
    .groups = "drop")|> 
  separate_wider_delim(
    population,
    delim = "_",
    names = c("latitude", "longitude"),
    cols_remove = FALSE) |> 
  mutate(
    latitude  = as.numeric(latitude),
    longitude = as.numeric(longitude) ) |> 
  arrange(latitude, longitude) |> 
  mutate(population = factor( population,levels = unique(population)))

map_points <- df_filtered2 |> 
  select(species, site_code, total, population, latitude, longitude) |> 
  distinct(species, site_code, .keep_all = TRUE)

plot_groups <- data_pop_year |> 
  distinct(species)

# facet plot 
data_pop_year |> 
  #filter(species == "Acanthaluteres vittiger") |> 
  ggplot(aes(x = year, y = population, fill = n_individuals)) +
  geom_tile(colour = "white", linewidth = 0.2) +
 scale_fill_paletteer_c("ggthemes::Sunset-Sunrise Diverging")+
   facet_wrap(~ species,
             labeller = labeller(species = label_wrap_gen(width = 10)), 
             ncol = 9, scales = "free") +
  theme_bw() +
  labs(
    x = "Year",
    y = "Population",
    fill = "Individuals")+
  theme(axis.text.y = element_blank())

pop_palette <- paletteer_d("MetBrewer::Redon")

# loop through species
pwalk(plot_groups, function(species) {
  
  hmap <- data_pop_year |> 
 filter(species == !!species)
  
  map_data <- map_points |>
    filter(species == !!species)
  
 p_hmap <-  ggplot(hmap, aes(x = year, y = population, fill = n_individuals)) +
  geom_tile(colour = "white", linewidth = 0.2) +
 scale_fill_paletteer_c("ggthemes::Sunset-Sunrise Diverging")+
  theme_bw() +
  labs(title = paste0("<i>", species, "</i>"),
    x = "Year",
    y = "Population",
    fill = "Individuals")+
   theme(plot.title = ggtext::element_markdown(size = 13))
  
  p_map <- ggplot() +
    geom_sf(data = aus, fill = "grey85", colour = "grey60", linewidth = 0.2) +
    geom_sf(data = aus_states, fill = NA, colour = "grey65", linewidth = 0.25) +
    geom_point(
      data = map_data,
      aes(x = longitude, y = latitude, colour = population),
      size = 0.8,
      alpha = 0.8, 
      shape = 16) +
    coord_sf(
      xlim = xlim,
      ylim = ylim,
      expand = FALSE) +
    scale_colour_manual(values = pop_palette) +
    scale_y_continuous(breaks = seq(-45.5, -10.5, by = 1),
                       labels = function(x) ifelse(seq_along(x) %% 4 == 1, x, "")) +
  scale_x_continuous(breaks = seq(135.5, 160.5, by = 1),
                     labels = function(x) ifelse(seq_along(x) %% 5 == 1, x, "")) +
    labs(
       x = expression("Longitude ("*degree*"E)"),
        y = expression("Latitude ("*degree*"S)")) +
    theme_bw() +
    theme(
      axis.ticks = element_line(linewidth = 0.2),
      axis.text = element_text(size = 6),
      axis.title = element_blank(),
      strip.text = element_text(size = 8),
      legend.position = "none")
  
  p <- p_hmap + p_map +
    plot_layout(widths = c(2, 1))
  
  file_name <- paste(species) |> 
    str_replace_all("[^A-Za-z0-9]+", "_")
  
  ggsave(
    filename = paste0(
      "02_pipeline/01_data_cleaning/plots_species_hmap/",
      file_name,
      ".png"),
    plot = p,
    width = 12,
    height = 7,
    dpi = 600 )
})

```

```{r final choice species}

# find the species only found on lord howe island
lhi_only <- df_filtered2 |> 
  group_by(species) |> 
  summarise(
    only_lhi = all(location == "Lord Howe Island"),
    .groups = "drop"
  ) |> 
  filter(only_lhi) |> 
  pull(species)

# separate them out
df_isolated <- df_filtered2 |> 
  filter(species %in% lhi_only)

# separate out all other species
df_else <- df_filtered2 |> 
  filter(!species %in% lhi_only)

# summarise lhi only species
df_isolated_sum <- df_isolated |> 
  filter(year!=2026) |> 
  group_by(species) |> 
  summarise(obs = sum(total), 
            n_years = n_distinct(year))

# summarise all other species
df_else_sum <- df_else |> 
  filter(year!=2026) |> 
  group_by(species) |> 
  mutate(n_pop = n_distinct(population)) |> 
  filter(n_pop >=3) |> 
  ungroup() |> 
  group_by(species, population) |> 
  summarise(obs = sum(total), 
            n_years = n_distinct(year))

# make species list
sp_list <- c("Acanthaluteres vittiger", "Achoerodus viridis", "Aplodactylus lophodon", "Atypichthys strigatus", "Caesioperca rasor", "Chromis hypsilepis", "Dinolestes lewini", "Enoplosus armatus", "Hypoplectrodes maccullochi", "Morwong fuscus", "Notolabrus fucicola", "Notolabrus gymnogenis", "Notolabrus tetricus", "Olisthops cyanomelas", "Ophthalmolepis lineolatus", "Ostorhinchus limenus", "Parma microlepis", "Parma unifasciata", "Parupeneus spilurus", "Pempheris affinis", "Pempheris compressa", "Pempheris multiradiata", "Pictilabrus laticlavius", "Prionurus microlepidotus","Schuettea scalaripinnis", "Scorpis aequipinnis", "Scorpis lineolata", "Thalassoma lunare", "Upeneichthys lineatus", "Neoodax balteatus", "Trachinops caudimaculatus", "Trachinops taeniatus")

# lhi only species list
sp_list_isolated <- c("Pseudolabrus luculentus", "Neoglyphidodon polyacanthus", "Ostorhinchus norfolcensis", "Stegastes gascoynei", "Stegastes fasciolatus", "Prionurus maculatus", "Parma polylepis", "Kyphosus sectatrix", "Chaetodon tricinctus", "Ostorhinchus flavus", "Kyphosus bigibbus", "Anampses elegans")

# make final dfs 
df_final <- df_filtered2 |> 
  filter(species %in% sp_list)

df_final_iso <- df_filtered2 |> 
  filter(species %in% sp_list_isolated)

df_total <- bind_rows(df_final, df_final_iso)

# write_csv(df_final, "0_data/02_data_clean/rls_main_clean.csv")
# write_csv(df_final_iso, "0_data/02_data_clean/rls_lhi_clean.csv")
# write_csv(df_total, "0_data/02_data_clean/rls_combo_clean.csv")
```

```{r size frequency map and heat map plots}
# set up data 
d_points <- df_total |> 
  select(area, species, year, size_class, total) 

d_lines <- df_total |> 
  group_by(area, species, year, size_class) |> 
  summarise(
    n_obs_size = sum(total, na.rm = TRUE),
    .groups = "drop") 

plot_groups <- d_points |> 
  distinct(species, area)

map_points <- df_total |> 
  select(species, site_code, total, population, latitude, longitude) |> 
  distinct(species, site_code, .keep_all = TRUE)

data_pop_year <- df_total |> 
  group_by(species, population, year) |> 
  summarise(
    n_individuals = sum(total, na.rm = TRUE),
    n_size_classes = n_distinct(size_class),
    .groups = "drop")|> 
  separate_wider_delim(
    population,
    delim = "_",
    names = c("latitude", "longitude"),
    cols_remove = FALSE) |> 
  mutate(
    latitude  = as.numeric(latitude),
    longitude = as.numeric(longitude) ) |> 
  arrange(latitude, longitude) |> 
  mutate(population = factor( population,levels = unique(population)))


# loop through species
pwalk(plot_groups, function(species, area) {
  
  point_data <- d_points |> 
    filter(species == !!species, area == !!area)
  
  line_data <- d_lines |> 
    filter(species == !!species, area == !!area)
  
  map_data <- map_points |>
    filter(species == !!species)
  
  hmap <- data_pop_year |> 
 filter(species == !!species)
  
  point_max <- max(point_data$total, na.rm = TRUE)
  line_max  <- max(line_data$n_obs_size, na.rm = TRUE)
  
  scale_factor <- point_max / line_max
  
  pop_palette <- paletteer_d("MetBrewer::Redon")
 
 p_hmap <-  ggplot(hmap, aes(x = year, y = population, fill = n_individuals)) +
  geom_tile(colour = "white", linewidth = 0.2) +
 scale_fill_paletteer_c("ggthemes::Sunset-Sunrise Diverging")+
  theme_bw() +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 5))+
  labs(
    x = "Year",
    y = "Population",
    fill = "Individuals")+
   theme(axis.ticks = element_line(linewidth = 0.2),
      axis.text = element_text(size = 6),
      axis.title = element_text(size = 7), 
      legend.title = element_text(size = 7), 
      legend.text = element_text(size = 6))
  
  p_years <- ggplot(point_data, aes(x = size_class)) +
    geom_point(aes(y = total), size = 0.7, alpha = 0.8, shape = 16) +
    geom_line(
      data = line_data,
      aes(y = n_obs_size * scale_factor, group = 1),
      colour = "darkcyan",
      linewidth = 0.4) +
    scale_y_continuous(
      name = "Total per survey",
      sec.axis = sec_axis( ~ . / scale_factor, name = "Cumulative total") ) +
    facet_wrap(~ year, scales = "free_y", ncol = 5) +
    labs(title = paste0("<i>", species, "</i> | ", area),
      x = "Size class" ) +
    theme_bw() +
    theme(plot.title = ggtext::element_markdown(size = 13),
      axis.text = element_text(size = 6),
      strip.text = element_text(size = 8), 
      axis.title = element_text(size = 12), 
      panel.grid.minor = element_blank())
  
  p_map <- ggplot() +
    geom_sf(data = aus, fill = "grey85", colour = "grey60", linewidth = 0.2) +
    geom_sf(data = aus_states, fill = NA, colour = "grey65", linewidth = 0.25) +
    geom_point(
      data = map_data,
      aes(x = longitude, y = latitude, colour = population),
      size = 0.8,
      alpha = 0.8, 
      shape = 16) +
    coord_sf(
      xlim = xlim,
      ylim = ylim,
      expand = FALSE) +
    scale_colour_manual(values = pop_palette) +
    scale_y_continuous(breaks = seq(-45.5, -10.5, by = 1),
                       labels = function(x) ifelse(seq_along(x) %% 4 == 1, x, "")) +
  scale_x_continuous(breaks = seq(135.5, 160.5, by = 1),
                     labels = function(x) ifelse(seq_along(x) %% 5 == 1, x, "")) +
    labs(
       x = expression("Longitude ("*degree*"E)"),
        y = expression("Latitude ("*degree*"S)")) +
    theme_bw() +
    theme(
      axis.ticks = element_line(linewidth = 0.2),
      axis.text = element_text(size = 6),
      axis.title = element_text(size = 7),
      strip.text = element_text(size = 8),
      legend.position = "none")
  
  p_side <- p_map/p_hmap +plot_layout(heights = c(1.8, 1))
  
  p <- p_years + p_side +
    plot_layout(widths = c(4, 0.85))
  
  file_name <- paste(species, area, sep = "_") |> 
    str_replace_all("[^A-Za-z0-9]+", "_")
  
  ggsave(
    filename = paste0(
      "02_pipeline/01_data_cleaning/plots_species_area_combo/",
      file_name,
      ".png"),
    plot = p,
    width = 12,
    height = 7,
    dpi = 600 )
})
```

