---
title: "Chapter 1 Model: Fit (with random effects)"
date: 2024-11-16
author:
  - name: Shane A. Richards
    id: sar
    orcid: 0000-0002-9638-5827
    email: shane.richards@utas.edu.au
    affiliation: 
      - name: School of Natural Sciences, University of Tasmania
        city: Hobart
        state: Tas, Australia
        url: https://discover.utas.edu.au/shane.richards
abstract: > 
  An alterate presentation of the model presented in Chapter 1. The model contain randoms effects at the site and year level, and the model has been fit to a data set of lengths for the species *Notolabrus tetricus*.
keywords:
  - Size-frequency distribution
  - Statistical ecology
format:
  html:
    self-contained: true
editor: visual
toc: true
code-overflow: wrap
theme: cerulean
---

# Methods

## Population model

To generate fish length distributions, we use an approach similar to the length-based spawning potential ratio (LB-SPR) method (Hordyk et al. 2015a). Fish are assumed to follow Bertalanffy growth, where the length of a fish of age $a$ is described by the equation: $$L(a) = L_\infty \big( 1 - \exp(-K(a-a_0)) \big ).$$ This equation incorporates growth rate ($K$), asymptotic length ($L_\infty$), and ($a_0 \leq 0$), the hypothetical age at which length is zero.

Fish exhibit a constant mortality rate ($M$), and we track them up to the age where they have a probability ($\varepsilon$) of being alive. This corresponds to age $a^* = -\ln(\varepsilon) / M$. For simplicity, we assume constant recruitment and that the fish population is at equilibrium. This implies that the age distribution of the population follows a negative exponential distribution. Additionally, variation in fish lengths arises from between-individual differences in $L_\infty$, which follows a normal distribution with mean $\bar{L}_\infty$ and a coefficient of variation, $c$.

The resulting length-distribution is approximated by discretising age into $X$ intervals of width $\Delta_x = a^*/X$. The probability of being alive at age $a_x = x \Delta_x$ is $S_x = \exp(-Ma_x)$. The expected length at age $a_x$ is $\bar{L}_x = L(a_x | \bar{L}_\infty, K, a_0)$. Here, we set $a_0 = 0$, which results in $\bar{L}_x$ being a function of $\bar{L}_\infty$ and the ratio $M/K$. The negative-exponential age-distribution is transformed into a length-distribution by discretising length into $J$ intervals, where the $j$^th^ interval is bounded by lengths $L_{j-1}$ and $L_j$. The relative abundance of fish in length interval $j$ is $$\tilde{f}_j = \begin{cases}
  \sum_{x=1}^X S_x \Phi \Big ( \frac{L_1-\bar{L}_x}{c\bar{L}_x} \Big ), & 
    \text{if } j = 1;  \\
  \sum_{x=1}^X S_x \Big ( \Phi \Big ( \frac{L_j-\bar{L}_x}{c\bar{L}_x} \Big ) -
    \Phi \Big ( \frac{L_{j-1}-\bar{L}_x}{c\bar{L}_x} \Big ) \Big ), & 
    \text{if } 1 < j < J;  \\
  \sum_{x=1}^X S_x \Big ( 1 - \Phi\Big ( \frac{L_{J}-\bar{L}_x}{c\bar{L}_x} \Big ) \Big ), & 
    \text{if } j = J;
\end{cases}$$

where $\Phi$ is the cumulative probability density function of the standard normal distribution. The proportion of fish in length interval $j$ is $f_j = \tilde{f}_j / \sum_{j=1}^J \tilde{f}_j$.

The proportion of fish observed in length-interval $j$, denoted $g_j$, will differ to the true proportion of fish in the interval ($f_j$) if the probability of observing fish is affected by their length. Here, we assume that smaller fish are more difficult to observe, such that fish of lengths $L_{50}$ and $L_{95}$ have a 50% and 95% chance of being observed, respectively. Specifically, the probability a fish in length interval $j$ is observed is given by $$s_j = \bigg ( 1 + \exp \bigg ( -\frac{\ln19}{L_{95} - L_{50}} \bigg ( \frac{ L_{j-1}+L_j }{2} - L_{50} \bigg ) \bigg ) \bigg )^{-1}.$$ Given these length-dependent probabilities of observation, an estimate of the proportion of fish observed in length interval $j$ is $$g_j = \frac{s_j\tilde{f}_j} {\sum_{j=1}^J s_j\tilde{f}_j}.$$

## Random effects

Copilot said:
The $M/K$ ratio for a given location and year is influenced by two fixed parameters: the mean sea-surface temperature for that year and whether the location is a marine protected area. Additionally, the $M/K$ ratio is adjusted by two random effects. The first random effect is linked to the location, with eight distinct locations considered. The second random effect is associated with each year-location combination, totaling 135 combinations.

## Data Visualisation

```{r}
#| message: false
#| warning: false

library(tidyverse) # data frame functionality
library(cowplot)   # multi-plot with ggplot
library(rstan)     # access Stan programming language
	
rm(list = ls()) # clear memory

# read in size data
df_raw <- read_csv("Notolabrus_tetricus.csv", show_col_types = FALSE) |>
  arrange(Loc, year, SizeClass)

glimpse(df_raw)
```

```{r}
#| fig-width: 9
#| fig-asp: 0.5
#| fig-align: center

pl_sst <- ggplot(df_raw) + 
  geom_line(aes(x = SizeClass, y = f, group = rep, color = Esst)) + 
  scale_color_gradient2(low = "blue", mid = "plum", high = "red", 
    midpoint = 17.5) +
  labs(x = "Size class (cm)", y = "Fraction", color = "Temp.") + 
	facet_wrap( ~ Loc, ncol = 4) + 
  theme_bw() + theme(panel.grid = element_blank())

pl_year <- ggplot(df_raw) + 
  geom_line(aes(x = SizeClass, y = f, group = rep, color = year)) + 
  scale_color_gradient2(low = "blue", mid = "plum", high = "red", 
    midpoint = 2008) +
  labs(x = "Size class (cm)", y = "Fraction", color = "Year") + 
	facet_wrap( ~ Loc, ncol = 4) + 
  theme_bw() + theme(panel.grid = element_blank())

plot_grid(pl_sst, pl_year, labels = c("a", "b"))
```

# Modelling

## Data Preparation

```{r}
# prep data for Stan fitting
X   <- 60 # age-classes
eps <- 0.01
S   <- exp(log(eps)*(1:X)/X) 

size_classes <- sort(unique(df_raw$SizeClass))
SC_max <- max(size_classes)
df_raw$SizeClass <- factor(df_raw$SizeClass, levels = size_classes)

# get concensus mpa status for each location
df_mpa <- df_raw |>
  select(Loc, year, mpa) |>
  unique() |>
  arrange(Loc, year)

df_mpa_grp <- df_mpa |>
  summarise(.by = "Loc", mpa = round(mean(mpa),0))

df_fit <- df_raw |>
  arrange(SizeClass) |>
  select(Loc, year, rep, Esst, SizeClass, fish_count) |>
	left_join(df_mpa_grp, by = "Loc") |>
  pivot_wider(names_from = SizeClass, values_from = fish_count) |>
  arrange(Loc, year)

df_fit <- df_fit %>%
  mutate(across(everything(), ~ replace_na(.x, 0))) # replace all NAs with 0

N <- nrow(df_fit) # location x year data sets
J <- length(size_classes) # number of size classes
L <- rep(0, J+1)  # length-class boundaries
Lmax <- SC_max + 10 # maximum size
for (j in 2:J) {
  L[j] <- 0.5*(size_classes[j-1] + size_classes[j])
}
L[J+1] <- Lmax # final length bin

# z-transform sea-surface temperatures
Esst_E  <- mean(df_fit$Esst)
Esst_sd <- sd(df_fit$Esst)
Esst_z  <- (df_fit$Esst - Esst_E) / Esst_sd 

# extract fish counts
m_n <- as.matrix(df_fit[ , 6:(5+J)]) # matrix of counts
nT <- apply(m_n, 1, sum) # total counts per set

r <- as.integer(factor(df_fit$Loc)) # location (integer)
R <- max(r)                         # number of locations

# create a list of data to pass to Stan
stan_dat <- list(
  N     = N,        # number of sets
	R     = R,        # number of locations (random effect)
  J     = J,        # number of size classes
  X     = X,        # age classes
  lnEps = log(eps), # log of epsilon
  S     = S,        # age-dependent survival
  Lmid  = size_classes, # middle of length box
  L     = L,        # length boundaries
	r     = r,        # location index
	mpa   = df_fit$mpa, # mpa status
  zT    = Esst_z,   # z-transformed sea-surface temperatures
  nT    = nT,       # total number of fish per set
  y     = m_n       # fish counts [set, sizeclass]
)

glimpse(stan_dat)
```

## Stan Model

```{stan output.var = "distribution_model"}
data {
  int<lower=1>              N;      // number of sets
  int<lower=1>              R;      // number of locations
  int<lower=1>              J;      // number of size classes
  int<lower=1>              X;      // age-classes
  real                  lnEps;      // ln(epsilon)
  real<lower=0.001,upper=1> S[X];   // probability alive
  real<lower=0>          Lmid[J];   // length boundaries
  real<lower=0>             L[J+1]; // length boundaries
  int<lower=1,upper=R>      r[N];   // location index
  int<lower=0,upper=1>    mpa[N];   // mpa status: 0 = no, 1 = yes
  real                     zT[N];   // z-transformed sea-surface temperatures
  int<lower=1>             nT[N];   // total number of fish per set
  int<lower=0>              y[N,J]; // fish counts [set, sizeclass]
}

parameters {
  real<lower= 1.0,upper=50.00> phi;        // multinomial overdispersion parameter
  real<lower=30.00,upper=55.00> ELInf;     // expected maximum size 
  real<lower= 0.50,upper= 2.50> MpK;       // typical M / K across sites and years
  real<lower=-0.50,upper= 0.50> beta_z;    // effect of temperature on M / K
  real<lower=-0.50,upper= 0.50> beta_mpa;  // effect of mpa on M / K
  real<lower= 0.00,upper=25.00> L50;       // length when 50% observed                     
  real<lower= 1.00,upper=20.00> DL95;      // length when 95% observed= L50+DL95                   
  real<lower= 0.05,upper= 0.30> cv;        // coefficient of variation in LInf
  real<lower=0.1,upper=1.00>    sigma_Loc; // between location variation                 
  real<lower=0.1,upper=1.00>    sigma_N;   // between year:location variation                 
  real MPK_Loc_RE[R];   // random deviations around overall mean for all locations
  real MPK_N_RE[N];   // random deviations around overall mean for all locations
}

transformed parameters{
}

model {
  vector[J] g;     // probability in length bin j
  vector[J] alpha; // multinomial parameter for length bin j
  vector[J] s;     // probability observe
  vector[X] Lbar;  // expected length at age
  
  array[J, X] real z; // z-transformed length

  real sum1;
  real sum2;
  real alpha0;
  real MpKi; // M / K for specific i
  
  // set priors 
  phi      ~ exponential(0.1);
  ELInf    ~ normal(40.0, 10.0);
  MpK      ~ normal( 1.5,  0.5);
  beta_z   ~ normal( 0.0,  0.2);
  beta_mpa ~ normal( 0.0,  0.2);
  L50      ~ normal(15.0,  5.0);
  DL95     ~ normal(10.0,  5.0);
  cv       ~ normal( 0.1,  0.05);
  sigma_Loc ~ normal(0.0,  0.5);
  sigma_N   ~ normal(0.0,  0.5);

  MPK_Loc_RE ~ normal(0.0, sigma_Loc); // random effect deviations (vectorised) 
  MPK_N_RE   ~ normal(0.0, sigma_N);   
  
  alpha0 = 1.0 / phi; // alpha sum
  
  for (j in 1:J) { // probability observed when in length bin j
    s[j] = 1.0 / (1 + exp(-(log(19)/DL95)*(Lmid[j]-L50)));
  }

  for (i in 1:N) { // for each set of counts
    // set specific M /K
    MpKi = exp(log(MpK) + beta_z*zT[i] + beta_mpa*mpa[i] + 
      MPK_Loc_RE[r[i]] +  MPK_N_RE[i]); 
  
    for (x in 1:X) { // expected length at age
      Lbar[x] = ELInf*(1.0 - exp(lnEps*x/(X*MpKi)));
    }
    
    for (x in 1:X) {
      for (j in 1:J) { // z-transformed length-at-age
        z[j,x] = (L[j+1] - Lbar[x])/(cv*Lbar[x]);
      }
    }
    
    // calculate g_j for this observation i
    
    sum2 = 0.0; // first length box
    for (x in 1:X) {
      sum2 += S[x]*normal_cdf(z[1,x], 0, 1);
    }
    g[1] = s[1]*sum2;
    sum1 = g[1];
    
    for (j in 2:(J-1)) { // intermediate length boxes
      sum2 = 0.0; 
      for (x in 1:X) {
        sum2 += S[x]*(normal_cdf(z[j,x], 0, 1) - normal_cdf(z[j-1,x], 0, 1));
      }
      g[j] = s[j]*sum2;
      sum1 += g[j];
    }
  
    sum2 = 0.0; // last length box
    for (x in 1:X) {
      sum2 += S[x]*(1.0 - normal_cdf(z[J,x], 0, 1));
    }
    g[J] = s[J]*sum2;
    sum1 += g[J];
    
    // calculate alpha
    for (j in 1:J) {
       g[j] = g[j] / sum1; // normalise
       alpha[j] = g[j] / alpha0; // scale probabilities to multinomial params
    }
    
    // calculate log-likelihood
    sum1 = lgamma(alpha0) + lgamma(nT[i] + 1) - lgamma(nT[i] + alpha0);
    for (j in 1:J) { // for each sub-period
      sum1 += lgamma(y[i,j] + alpha[j]) - 
        lgamma(alpha[j]) - lgamma(y[i,j] + 1); 
    }
    
    target += sum1; // add log-likelihood term associated with set of obs
  }
}

generated quantities {
}
```

## Model Fitting

```{r}
# fit the Stan model
fit_1 <- rstan::sampling(object = distribution_model, data = stan_dat,
  iter = 300, warmup = 100, chains = 1, refresh = 20)
```

```{r}
# vector of model parameter names
params <- c("ELInf", "MpK", "L50", "DL95", "cv", "beta_z", "beta_mpa",
  "sigma_Loc", "sigma_N", "phi")

# check for posterior convergence
traceplot(fit_1, pars = params, inc_warmup = TRUE, ncol = 4)
```

```{r}
# show the posterior parameter statistics
print(fit_1, probs = c(0.055,0.5,0.945), digits=3) # 89% credible intervals
```

```{r}
l_pars <- extract(fit_1) # extract posterior samples
```

# Predictions

## M/K and $L_\infty$

```{r}
df_MpK <- tibble(
    Esst = seq(from = min(df_fit$Esst), to = max(df_fit$Esst), 
      length.out = 20)
  ) |>
  mutate(
  	zT = (Esst - Esst_E) / Esst_sd, 
  	MpK_mdn = 0, MpK_lwr = 0, MpK_upr = 0) # median, lower

for (i in 1:nrow(df_MpK)) {
  MpK <- exp(log(l_pars$MpK) + l_pars$beta_z*df_MpK$zT[i])
  df_MpK$MpK_mdn[i] <- median(MpK)
  df_MpK$MpK_lwr[i] <- quantile(MpK, probs = 0.05)
  df_MpK$MpK_upr[i] <- quantile(MpK, probs = 0.95)
}

df_s <- tibble(
    Lmid = size_classes
  ) |>
  mutate(
  	s_mdn = 0, s_lwr = 0, s_upr = 0) # median, lower

for (i in 1:nrow(df_s)) {
  s <- 1.0 / (1 + exp(-(log(19)/l_pars$DL95)*(df_s$Lmid[i]-l_pars$L50)));

  df_s$s_mdn[i] <- median(s)
  df_s$s_lwr[i] <- quantile(s, probs = 0.05)
  df_s$s_upr[i] <- quantile(s, probs = 0.95)
}

df_ELInf <- tibble(ELInf = l_pars$ELInf)

df_cv <- tibble(cv = l_pars$cv)
```

```{r}
#| fig-width: 8
#| fig-asp: 0.65
#| fig-align: center
#| fig-cap: Posterior predictions. (a) Effect of sea surface temperature on M/K at unprotected locations. (b) Length-dependent probability of detection. (c) Mean maximum length. (d) Coefficient of variation of fish lengths for a given age. Shaded regions depict 90% credible intervals. 
#| label: fig-PostPredParam

pl_MpK <- ggplot(df_MpK) + 
  geom_ribbon(aes(x = Esst, ymin = MpK_lwr, ymax = MpK_upr), fill = "#b2e2e2") +
  geom_line(aes(x = Esst, y = MpK_mdn), color = "#238b45") +
  labs(x = "Mean sea surface temperature (C)", y = "M/K") +  
  theme_bw()

pl_s <- ggplot(df_s) + 
  geom_vline(xintercept = L, linetype = "dashed", color = "grey") +
  geom_ribbon(aes(x = Lmid, ymin = s_lwr, ymax = s_upr), fill = "#b2e2e2") +
  geom_line(aes(x = Lmid, y = s_mdn), color = "#238b45") +
  labs(x = "Length (cm)", y = "Probability observe") +  
  theme_bw() + theme(panel.grid = element_blank())

pl_ELInf <- ggplot(df_ELInf) + 
  geom_histogram(aes(x = ELInf, y = after_stat(density)), 
    binwidth = 0.5, fill = "#b2e2e2", color = "#238b45") +
  labs(x = "Mean maximum length (cm)", y = "Fraction") +  
  theme_bw()

pl_cv <- ggplot(df_cv) + 
  geom_histogram(aes(x = cv, y = after_stat(density)), 
    binwidth = 0.01, fill = "#b2e2e2", color = "#238b45") +
  labs(x = "Coefficient of variation", y = "Fraction") +  
  theme_bw()

plot_grid(pl_MpK, pl_s, pl_ELInf, pl_cv, labels = c("a", "b", "c", "d"))
```

```{r}
#| fig-width: 6
#| fig-asp: 0.8
#| fig-align: center
#| fig-cap: Posterior parameter correlations.
#| label: fig-PostCorr

pairs(as.data.frame(l_pars) |> select(ELInf, MpK, cv, beta_z, beta_mpa))
```

## Size Distribution

```{r}
REPS  <- length(l_pars$phi)
eps   <- 0.005 # upper survival
lnEps <- log(eps)

Essts <- c(13, 16, 19)
zT <- (Essts - Esst_E) / Esst_sd
```

```{r}
# predict for a typical location
N_predict <- length(Essts)

m_g_mdn <- matrix(data = NA, nrow = N_predict, ncol = J)
m_g_lwr <- matrix(data = NA, nrow = N_predict, ncol = J)
m_g_upr <- matrix(data = NA, nrow = N_predict, ncol = J)

Lbar <- matrix(data = NA, nrow = REPS, ncol = X)
z    <- array(data = NA, dim = c(REPS, J, X))

alpha0 = 1.0 / l_pars$phi # alpha sum
s   <- matrix(data = NA, nrow = REPS, ncol = J)
for (j in 1:J) { # probability observed when in length bin j
  s[ ,j] = 1.0 / (1 + exp(-(log(19)/l_pars$DL95)*(size_classes[j]-l_pars$L50)))
}

for (n in 1:N_predict) {
  g    <- matrix(data = 0, nrow = REPS, ncol = J)
  MpKi <- exp(log(l_pars$MpK) + l_pars$beta_z*zT[n]) # set specific M /K
  
  for (x in 1:X) { # expected length at age
    Lbar[ ,x] <- l_pars$ELInf*(1.0 - exp(lnEps*x/(X*MpKi)))
  }
    
  for (x in 1:X) {
    for (j in 1:J) { # z-transformed length-at-age
      z[ ,j,x] = (L[j+1] - Lbar[ ,x])/(l_pars$cv*Lbar[ ,x])
    }
  }
    
  # calculate g_j for this prediction
    
  sum1 <- rep(0, REPS)
  
  sum2 <- rep(0, REPS)
  for (x in 1:X) {
    sum2 <- sum2 + S[x]*pnorm(z[ ,1,x])
  }
  g[ ,1] = s[ ,1]*sum2
  sum1 = g[ ,1]
    
  for (j in 2:(J-1)) { # intermediate length boxes
    sum2 <- rep(0, REPS) 
    for (x in 1:X) {
      sum2 <- sum2 +  S[x]*(pnorm(z[ ,j,x]) - pnorm(z[ ,j-1,x]))
    }
    g[ ,j] = s[ ,j]*sum2
    sum1 <- sum1 + g[ ,j]
  }
  
  sum2 <- rep(0, REPS) # last length box
  for (x in 1:X) {
    sum2 <- sum2 + S[x]*(1.0 - pnorm(z[ ,J,x]))
  }
  g[ ,J] = s[ ,J]*sum2
  sum1 <- sum1 + g[ ,J]
    
  # calculate g
  for (j in 1:J) {
    g[ ,j] = g[ ,j] / sum1 # normalise
  }
	
  m_g_mdn[n, ] <- apply(X = g, MARGIN = 2, FUN = median)
  m_g_lwr[n, ] <- apply(X = g, MARGIN = 2, FUN = quantile, probs = 0.05)
  m_g_upr[n, ] <- apply(X = g, MARGIN = 2, FUN = quantile, probs = 0.95)
}

df_g <- tibble(
  Lmid = size_classes,
  Esst = Essts[1],
  g_lwr = m_g_lwr[1, ],
  g_mdn = m_g_mdn[1, ],
  g_upr = m_g_upr[1, ])

for (n in 2:N_predict) {
  df_g <- df_g |>
    rbind(tibble(
      Lmid = size_classes,
      Esst = Essts[n],
      g_lwr = m_g_lwr[n, ],
      g_mdn = m_g_mdn[n, ],
      g_upr = m_g_upr[n, ]))
}

df_g$EsstF <- factor(df_g$Esst)
```

```{r}
# predict for the observed locations
Locs <- sort(unique(df_fit$Loc))
df_gLoc <- expand_grid(Loc = Locs, Esst = Essts) |>
  left_join(df_mpa_grp, by = "Loc")
df_gLoc$Loc <- factor(df_gLoc$Loc)
df_gLoc$iLoc <- as.integer(df_gLoc$Loc)
df_gLoc$iEsst <- as.integer(factor(df_gLoc$Esst))


N_predict <- nrow(df_gLoc)

m_gLoc_mdn <- matrix(data = NA, nrow = N_predict, ncol = J)
m_gLoc_lwr <- matrix(data = NA, nrow = N_predict, ncol = J)
m_gLoc_upr <- matrix(data = NA, nrow = N_predict, ncol = J)

Lbar <- matrix(data = NA, nrow = REPS, ncol = X)
z    <- array(data = NA, dim = c(REPS, J, X))

alpha0 = 1.0 / l_pars$phi # alpha sum
s   <- matrix(data = NA, nrow = REPS, ncol = J)
for (j in 1:J) { # probability observed when in length bin j
  s[ ,j] = 1.0 / (1 + exp(-(log(19)/l_pars$DL95)*(size_classes[j]-l_pars$L50)))
}

for (n in 1:N_predict) {
  g    <- matrix(data = 0, nrow = REPS, ncol = J)
  MpKi <- exp(log(l_pars$MpK) + 
    l_pars$beta_z*zT[df_gLoc$iEsst[n]] + l_pars$beta_mpa*df_gLoc$iEsst[n] +
    l_pars$MPK_Loc_RE[ ,df_gLoc$iLoc[n]]) # set specific M /K
  
  for (x in 1:X) { # expected length at age
    Lbar[ ,x] <- l_pars$ELInf*(1.0 - exp(lnEps*x/(X*MpKi)))
  }
    
  for (x in 1:X) {
    for (j in 1:J) { # z-transformed length-at-age
      z[ ,j,x] = (L[j+1] - Lbar[ ,x])/(l_pars$cv*Lbar[ ,x])
    }
  }
    
  # calculate g_j for this prediction
    
  sum1 <- rep(0, REPS)
  
  sum2 <- rep(0, REPS)
  for (x in 1:X) {
    sum2 <- sum2 + S[x]*pnorm(z[ ,1,x])
  }
  g[ ,1] = s[ ,1]*sum2
  sum1 = g[ ,1]
    
  for (j in 2:(J-1)) { # intermediate length boxes
    sum2 <- rep(0, REPS) 
    for (x in 1:X) {
      sum2 <- sum2 +  S[x]*(pnorm(z[ ,j,x]) - pnorm(z[ ,j-1,x]))
    }
    g[ ,j] = s[ ,j]*sum2
    sum1 <- sum1 + g[ ,j]
  }
  
  sum2 <- rep(0, REPS) # last length box
  for (x in 1:X) {
    sum2 <- sum2 + S[x]*(1.0 - pnorm(z[ ,J,x]))
  }
  g[ ,J] = s[ ,J]*sum2
  sum1 <- sum1 + g[ ,J]
    
  # calculate g
  for (j in 1:J) {
    g[ ,j] = g[ ,j] / sum1 # normalise
  }
	
  m_gLoc_mdn[n, ] <- apply(X = g, MARGIN = 2, FUN = median)
  m_gLoc_lwr[n, ] <- apply(X = g, MARGIN = 2, FUN = quantile, probs = 0.05)
  m_gLoc_upr[n, ] <- apply(X = g, MARGIN = 2, FUN = quantile, probs = 0.95)
}

df_gLocL <- tibble(
	Loc = df_gLoc$Loc[1],
  Lmid = size_classes,
  Esst = df_gLoc$Esst[1],
  g_lwr = m_gLoc_lwr[1, ],
  g_mdn = m_gLoc_mdn[1, ],
  g_upr = m_gLoc_upr[1, ])

for (n in 2:N_predict) {
  df_gLocL <- df_gLocL |>
    rbind(tibble(
    	Loc = df_gLoc$Loc[n],
      Lmid = size_classes,
      Esst = df_gLoc$Esst[n],
      g_lwr = m_gLoc_lwr[n, ],
      g_mdn = m_gLoc_mdn[n, ],
      g_upr = m_gLoc_upr[n, ]))
}

df_gLocL$EsstF <- factor(df_gLocL$Esst)
```

```{r}
# observed frequencies
m_f <- m_n
for(n in 1:N) {
  m_f[n, ] <- m_n[n, ] / nT[n]	
}

df_f <- tibble(rep = 1:N, Loc = df_fit$Loc, Esst = Esst_z*Esst_sd + Esst_E) |>
  cbind(as.data.frame(m_f)) |>
  pivot_longer(names_to = "Lmid", values_to = "f", 
    cols = all_of(4:(3+J)), names_transform = as.numeric)

df_f <- df_f |>
  mutate(
    EsstF = ifelse(Esst < 0.5*(Essts[1] + Essts[2]), Essts[1],
      ifelse(Esst > 0.5*(Essts[2] + Essts[3]), Essts[3], Essts[2])))
df_f$EsstF <- factor(df_f$EsstF)
```

```{r}
#| fig-width: 5
#| fig-asp: 0.45
#| fig-align: center
#| fig-cap: Predicted fish length distributions for three sea surface temperatures, for locations not an mpa. Shaded regions depict 90% credible intervals. 
#| label: fig-PostSST1

ggplot() + 
  geom_ribbon(data = df_g,
    aes(x = Lmid, ymin = g_lwr, ymax = g_upr, fill = EsstF)) +
  geom_line(data = df_g, aes(x = Lmid, y = g_mdn, color = EsstF)) +
  scale_color_manual(values = c("#0570b0", "#238443", "#cb181d")) +
  scale_fill_manual(values = c("#bdc9e1", "#c2e699", "#fcae91")) +
  labs(x = "Length (cm)", y = "Fraction", 
    color = "Temperature", fill = "Temperature") +  
  theme_bw() + theme(panel.grid = element_blank())
```

```{r}
#| fig-width: 5
#| fig-asp: 0.5
#| fig-align: center
#| fig-cap: Random deviations in log(M/K) associated with the eight locations. Error bars depict the 90% credible interval.
#| label: fig-Post_SST1_RE_Loc

df_Loc <- t(apply(X = l_pars$MPK_Loc_RE, MARGIN = 2, FUN = quantile, probs = c(0.05, 0.5, 0.95))) |>
  as_tibble()
names(df_Loc) <- c("lwr", "mdn", "upr")
df_Loc$Loc <- levels(factor(df_fit$Loc))

df_Loc <- df_Loc |>
  left_join(df_mpa_grp, by = "Loc")
df_Loc$mpa <- factor(df_Loc$mpa)

ggplot(df_Loc) + 
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_errorbarh(aes(xmin = lwr, xmax = upr, y = Loc), height = 0.2) +
  geom_point(aes(x = mdn, y = Loc, color = mpa)) +
	labs(x = "Random deviation", y = "Location", color = "Protected") +
  theme_bw() + theme(panel.grid = element_blank())
```

```{r}
df_N <- t(apply(X = l_pars$MPK_N_RE, MARGIN = 2, FUN = quantile, probs = c(0.05, 0.5, 0.95))) |>
  as_tibble()
names(df_N) <- c("lwr", "mdn", "upr")
df_N$Loc <- df_fit$Loc
df_N$year <- df_fit$year

df_N <- df_N |>
  left_join(df_mpa, by = c("Loc", "year"))
df_N$mpa <- factor(df_N$mpa)
```

```{r}
#| fig-width: 7
#| fig-asp: 0.6
#| fig-align: center
#| fig-cap: Random deviations in log(M/K) associated with the 135 location x year combinations. Error bars depict the 90% credible interval.
#| label: fig-Post_SST1_RE_N

ggplot(df_N) + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = lwr, ymax = upr, x = year), width = 0.2) +
  geom_point(aes(y = mdn, x = year, color = mpa)) +
	facet_wrap( ~ Loc, ncol = 4) +
	labs(x = "Year", y = "Random deviation") +
  theme_bw() + theme(panel.grid = element_blank())
```

```{r}
#| fig-width: 5
#| fig-asp: 0.6
#| fig-align: center
#| fig-cap: Random deviations in log(M/K) associated with the 135 location x year combinations. Error bars depict the 90% credible interval.
#| label: fig-Post_SST1_RE_N2

ggplot(df_N) + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(aes(y = mdn, x = year, color = Loc)) +
	labs(x = "Year", y = "Random deviation") +
  theme_bw() + theme(panel.grid = element_blank())
```

```{r}
#| fig-width: 7
#| fig-asp: 1
#| fig-align: center
#| fig-cap: Predicted fish length distributions for three sea surface temperatures in relation to observed length distributions at each sampling location. Shaded regions depict 90% credible intervals. 
#| label: fig-PostSST2

ggplot() + 
  geom_line(data = df_f, aes(x = Lmid, y = f, group = rep),
    color = "grey") + 
  geom_ribbon(data = df_gLocL,
    aes(x = Lmid, ymin = g_lwr, ymax = g_upr, fill = EsstF)) +
  geom_line(data = df_gLocL, aes(x = Lmid, y = g_mdn, color = EsstF)) +
  scale_color_manual(values = c("#0570b0", "#238443", "#cb181d")) +
  scale_fill_manual(values = c("#bdc9e1", "#c2e699", "#fcae91")) +
  facet_grid(Loc ~ EsstF) +
  labs(x = "Length (cm)", y = "Fraction",
    color = "Temperature", fill = "Temperature") +  
  theme_bw() + theme(panel.grid = element_blank())
```

# Conclusions

## Major

-   M/K increases with mean sea-surface temperatures (@fig-PostPredParam a)
    -   environment is getting worse for this species
    -   increased temperature is not translating to increased growth
        -   increased mortality?
-   Protected locations generally have a lower estimated M/K ($\beta_\text{mpa} < 0$)
-   Annual variation in relative changes in estimated M/K is consistent across Tasmania (@fig-Post_SST1_RE_N2)

## Minor

-   Wide bin widths are making observation selectivity tricky to estimate (@fig-PostPredParam b)
    -   how good is the numerical approximation that selectivity holds equally across the length bin?
-   fit wants to lower $\bar{L}_\infty$ and increase the coefficient of variation (CoV) (@fig-PostCorr)
    -   how strong to make priors on CoV (@fig-PostPredParam d)?
-   There is so much variation in the frequencies between length bins, that it cannot be all explained by the Dirichelet multinomial distribution
    -   $\phi$ is estimated to be high (> 20) and has little additional effect on fit when increased further
    -   suggests that the model is structurally inadequate in some way
        -   missing predictors?
